Novel lattice Boltzmann method for simulation of strongly shear thinning viscoelastic fluids

### 2024-02-20 NovelLBM.tex
derivation PTT sol

For both Poiseuille flow and shear flow the following holds.
\begin{equation}
	\vec{u} = \dot{\gamma}\left(y\right)\hat{e}_x
\end{equation}
With this the constitutive equation\todo{link} fo $\xi=0$ in steady state reduces to the following.
\begin{align}
	\exp\left(\frac{\epsilon\lambda}{\eta_\text{p}}\Tr\ma{\tau}\right)\tau_{xx} &= 2\lambda\dot{\gamma}\tau_{xy} \\
	\exp\left(\frac{\epsilon\lambda}{\eta_\text{p}}\Tr\ma{\tau}\right)\tau_{xy} &= \eta_\text{p}\dot{\gamma}
\end{align}
With $\Tr\ma{\tau} = \tau_{xy}$. In the three-dimensional case $y$ refers to the radial coordinate.

To account for the added viscous stress due to $\eta_\text{s}$, equilibrium is reached at lower polymer stresses. Therefore, Oliveiras equation (7) needs to be modified as follows.
	\begin{equation}
		\tau_{xy} = \partial_xp\frac{r}{2^j} - \eta_\text{s}\dot{\gamma}
	\end{equation}
	Where the second term is added due to the presence of the solvent viscosity. This term also propagates to Oliveiras equation (9) as follows.
	\begin{align}
		\dot{\gamma} &= f\left(\frac{2\lambda}{\eta_\text{p}}\left[\partial_xp\frac{r}{2^j} - \eta_\text{s}\dot{\gamma}\right]^2\right)\frac{1}{\eta_\text{p}}\left(\partial_xp\frac{r}{2^j}-\eta_\text{s}\dot{\gamma}\right)
		\intertext{This cannot be integrated as easily as without $\eta_\text{s}$, and is only solvable semi-analytically.}
		1 &= f\left(\frac{2\lambda}{\eta_\text{p}}\left[\partial_xp\frac{r}{2^j} - \eta_\text{s}\dot{\gamma}\right]^2\right)\frac{1}{\eta_\text{p}}\left(\partial_xp\frac{r}{2^j\dot{\gamma}}-\eta_\text{s}\right)
	\end{align}
	For the exponential version of PTT $f$ is defined as follows.
	\begin{equation}
		f\left(\Tr\ma{\tau}\right) = \exp\left(\frac{\epsilon\lambda}{\eta_\text{p}}\Tr\ma{\tau}\right)
	\end{equation}
	This can be inserted in the above equation.
	\begin{align}
		1 &= \exp\left(\frac{2\epsilon\lambda^2}{\eta_\text{p}^2}\left[\partial_xp\frac{r}{2^j} - \eta_\text{s}\dot{\gamma}\right]^2\right)\frac{1}{\eta_\text{p}}\left(\partial_xp\frac{r}{2^j\dot{\gamma}}-\eta_\text{s}\right) \\
		0 &= \frac{2\epsilon\lambda^2}{\eta_\text{p}^2}\left[\partial_xp\frac{r}{2^j} - \eta_\text{s}\dot{\gamma}\right]^2 - \ln\eta_\text{p} + \ln\left(\partial_xp\frac{r}{2^j\dot{\gamma}}-\eta_\text{s}\right) \\
		0 &= 2\epsilon\lambda^2\dot{\gamma}^2\left[\partial_xp\frac{r}{2^j\dot{\gamma}\eta_\text{p}} - \frac{\eta_\text{s}}{\eta_\text{p}}\right]^2 + \ln\left(\partial_xp\frac{r}{2^j\dot{\gamma}\eta_\text{p}}-\frac{\eta_\text{s}}{\eta_\text{p}}\right) \\
		0 &= 2\epsilon\lambda^2\dot{\gamma}^2\left[\partial_xp\frac{r}{2^j\dot{\gamma}\eta_\text{p}} - \frac{\eta_\text{s}}{\eta_\text{p}}\right]^2 + \ln\left(\partial_xp\frac{r}{2^j\dot{\gamma}\eta_\text{p}}-\frac{\eta_\text{s}}{\eta_\text{p}}\right)
		\intertext{This has the following form}
		0 &= ab^2 + \ln b
	\end{align}
	This can be solved using the Lambert W function as follows.
	\begin{align}
		b &= -\sqrt{\frac{W_0\left(2a\right)}{2a}}
		\intertext{Resolving for $r$ yields the following.}
		r &=  \frac{2^j\eta_\text{s}}{\partial_xp}\dot{\gamma} - \frac{\eta_\text{p}2^{j-1}}{\lambda\sqrt{\epsilon}\partial_xp}\sqrt{W_0\left(4\epsilon\lambda^2\dot{\gamma}^2\right)}
	\end{align}
	We notice, that this is the solution for a poiseuille flow with viscosity $\eta_\text{s}$ expanded by a term that applies a nonuniform scaling of the profile  along the $r$-axis towards higher radii. We simplify again using $c < 0$ and $d > 0$.
	\begin{equation}
		r = c\dot{\gamma} - R_\eta\frac{c}{d}\sqrt{W_0\left(d^2\dot{\gamma}^2\right)}
	\end{equation}
	Finally this function needs to be inverted. This needs to be done numerically. We calculate a range of $\dot{\gamma}$ from $0$ to $\frac{R}{c}$ with a small step interval. Here $R$ is the radius of the channel or pipe. This is guaranteed to contain all relevant radii. We flip the axis and integrate numerically using scipy.integrate.cumulative\_trapezoid\todo{cite}. This is very accurate as we conveniently picked the spacing of our now $x$-axis in a way, that the $y$-axis ($\dot{\gamma}$) varies little for each step. From the recovered curve $u$ is calculated at each required $r$ through linear interpolation.

>>>>

For both Poiseuille flow and shear flow the following holds.
\begin{equation}
	\vec{u} = \dot{\gamma}\left(y\right)\hat{e}_x
\end{equation}
With this the constitutive equation\todo{link} fo $\xi=0$ in steady state reduces to the following.
\begin{align}
	\exp\left(\frac{\epsilon\lambda}{\eta_\text{p}}\Tr\ma{\tau}\right)\tau_{xx} &= 2\lambda\dot{\gamma}\tau_{xy} \label{eq:PTTDerivation1}\\
	\exp\left(\frac{\epsilon\lambda}{\eta_\text{p}}\Tr\ma{\tau}\right)\tau_{xy} &= \eta_\text{p}\dot{\gamma} \label{eq:PTTDerivation2}
\end{align}
With $\Tr\ma{\tau} = \tau_{xy}$. In the three-dimensional case $y$ refers to the radial coordinate.
Eliminating $\tau_{xy}$ yields the following.
\begin{equation}
	\exp\left(2\frac{\epsilon\lambda}{\eta_\text{p}}\Tr\ma{\tau}\right)\tau_{xx} = 2\lambda\dot{\gamma}^2\eta_\text{p}
\end{equation}
This is solved by the Lambert $W$ function as follows.
\begin{equation}
	\tau_{xx} = \frac{\eta_\text{p}}{2\epsilon\lambda}W_0\left(4\epsilon\lambda^2\dot{\gamma}^2\right)
\end{equation}
With $\tau_{yy} = 0$ this yields the following.
\begin{equation}
	N_1 = \sigma_{xx} - \sigma_{yy} = \tau_{xx} = \frac{\eta_\text{p}}{2\epsilon\lambda}W_0\left(4\epsilon \mathrm{Wi}^2\right) %\label{eq:shearN}
\end{equation}
For the viscosity, we insert this into equation~(\ref{eq:PTTDerivation2}) and retrieve the following.
\begin{align}
	\eta &= \frac{\sigma_{xy}}{\dot{\gamma}} \\
	&= \frac{\tau_{xy} + \eta_\text{s}\dot{\gamma}}{\dot{\gamma}} \\
	&= \eta\left(\dot{\gamma}\right) = \frac{\eta_\text{p}}{\exp\left[0.5W_0\left(4\epsilon \mathrm{Wi}^2\right)\right]} + \eta_\text{s} %\label{eq:shearEta}
\end{align}
This gives the relevant solutions for the shear flow. For Poiseuille flow, we start with the Cauchy momentum equation (neglecting gravity), which reads as follows.
\begin{equation}
	\rho\frac{\Dif\vec{u}}{\Dif t} = -\nabla p + \nabla\cdot\sigma
\end{equation}
In steady state for this geometry, once integrated along $y$, this reduces to the following.
\begin{equation}
	\sigma_{xy} = \frac{\partial p}{\partial x}\frac{y}{2^j}
\end{equation}
Here $j$ is $0$ for the two-dimensional case and $1$ in the three-dimensional flow. The total stress experienced by the fluid is the one described by the constitutive equation plus the solvent contribution and thus the following holds.
\begin{equation}
	\tau_{xy} = \partial_xp\frac{r}{2^j} - \eta_\text{s}\dot{\gamma}
\end{equation}
Here we switched from y to writing $y$ to $r$ in order to clarify, that this is a radial coordinate being $0$ in the center of the channel for both cases.
Dividing equation~(\ref{eq:PTTDerivation1}) by equation~(\ref{eq:PTTDerivation2}) leads to the following relation.
\begin{equation}
	\tau_{xx} = \frac{2\lambda}{\eta_\text{p}}\tau_{xy}^2 =  \frac{2\lambda}{\eta_\text{p}}\left(\partial_xp\frac{r}{2^j} - \eta_\text{s}\dot{\gamma}\right)^2
\end{equation}
Inserting this back into equation~(\ref{eq:PTTDerivation2}) yields the following.
	\begin{align}
		\dot{\gamma} &= \exp\left(\frac{2\epsilon\lambda^2}{\eta_\text{p}^2}\left[\partial_xp\frac{r}{2^j} - \eta_\text{s}\dot{\gamma}\right]^2\right)\frac{1}{\eta_\text{p}}\left(\partial_xp\frac{r}{2^j}-\eta_\text{s}\dot{\gamma}\right)
		\intertext{This cannot be integrated as easily as without $\eta_\text{s}$, and is only solvable semi-analytically.}
		1 &= \exp\left(\frac{2\epsilon\lambda^2}{\eta_\text{p}^2}\left[\partial_xp\frac{r}{2^j} - \eta_\text{s}\dot{\gamma}\right]^2\right)\frac{1}{\eta_\text{p}}\left(\partial_xp\frac{r}{2^j\dot{\gamma}}-\eta_\text{s}\right) \\
		0 &= \frac{2\epsilon\lambda^2}{\eta_\text{p}^2}\left[\partial_xp\frac{r}{2^j} - \eta_\text{s}\dot{\gamma}\right]^2 - \ln\eta_\text{p} + \ln\left(\partial_xp\frac{r}{2^j\dot{\gamma}}-\eta_\text{s}\right) \\
		0 &= 2\epsilon\lambda^2\dot{\gamma}^2\left[\partial_xp\frac{r}{2^j\dot{\gamma}\eta_\text{p}} - \frac{\eta_\text{s}}{\eta_\text{p}}\right]^2 + \ln\left(\partial_xp\frac{r}{2^j\dot{\gamma}\eta_\text{p}}-\frac{\eta_\text{s}}{\eta_\text{p}}\right) \\
		0 &= 2\epsilon\lambda^2\dot{\gamma}^2\left[\partial_xp\frac{r}{2^j\dot{\gamma}\eta_\text{p}} - \frac{\eta_\text{s}}{\eta_\text{p}}\right]^2 + \ln\left(\partial_xp\frac{r}{2^j\dot{\gamma}\eta_\text{p}}-\frac{\eta_\text{s}}{\eta_\text{p}}\right)
		\intertext{This has the following form}
		0 &= ab^2 + \ln b
	\end{align}
	This can be solved using the Lambert $W$ function as follows.
	\begin{align}
		b &= -\sqrt{\frac{W_0\left(2a\right)}{2a}}
		\intertext{Resolving for $r$ yields the following.}
		r &=  \frac{2^j\eta_\text{s}}{\partial_xp}\dot{\gamma} - \frac{\eta_\text{p}2^{j-1}}{\lambda\sqrt{\epsilon}\partial_xp}\sqrt{W_0\left(4\epsilon\lambda^2\dot{\gamma}^2\right)}
	\end{align}
	We notice, that this is the solution for a poiseuille flow with viscosity $\eta_\text{s}$ expanded by a term that applies a nonuniform scaling of the profile  along the $r$-axis towards higher radii. We simplify again using $c < 0$ and $d > 0$.
	\begin{equation}
		r = c\dot{\gamma} - R_\eta\frac{c}{d}\sqrt{W_0\left(d^2\dot{\gamma}^2\right)}
	\end{equation}
	Finally this function needs to be inverted. This needs to be done numerically. We calculate a range of $\dot{\gamma}$ from $0$ to $\frac{R}{c}$ with a small step interval. Here $R$ is the radius of the channel or pipe. This is guaranteed to contain all relevant radii. We flip the axis and integrate numerically using scipy.integrate.cumulative\_trapezoid\todo{cite}. This is very accurate as we conveniently picked the spacing of our now $x$-axis in a way, that the $y$-axis ($\dot{\gamma}$) varies little for each step. From the recovered curve $u$ is calculated at each required $r$ through linear interpolation.
